---
title: Setting output args via render functions
runtime: shiny
output: html_document
---

This interactive Rmd document makes use of the `outputArgs`
argument now available to all Shiny `render` functions. To
give an example, this allows you to set arguments to
`imageOutput` through `renderImage`. This means that you
don't have to create a `ui` object just to be able to brush
over an image. Note that this only applies to snippets of
Shiny code during an interactive Rmd (and not to embedded
full apps -- the ones you need to call `shinyApp` to run).

## Brushing over an image (and storing the data)

```{r setup, echo=FALSE}
library(datasets)

sliderInput(label = 'foo', 'sample', min= 10, max= 500, value = 100)


generateImage = function(x) {
  outfile = tempfile(fileext = '.png')
  png(outfile)
  hist(x)
  dev.off()
  list(src = outfile)
}
```

```{r image}
renderImage({
  generateImage(runif(input$sample))
}, deleteFile = TRUE, 
   outputArgs = list(brush = brushOpts(id = "plot_brush"),
                     width = "1200px",
                     height = "400px")
)
```



Here is some of the brushing info sent to the server:
(brush over the image to change the data)

```{r brush info}
renderText({ 
  print(input$plot_brush)
  brush = input$plot_brush
  paste0("xmin: ", brush$xmin, "; ",
         "xmax: ", brush$xmax, "; ",
         "ymin: ", brush$ymin, "; ",
         "ymax: ", brush$ymax)
})
```

---

### Resizing a plot

```{r plot}
renderPlot({ 
  plot(cars) 
}, outputArgs = list(width = "75%", 
                     height = "250px")
)
```